---
output: cynkratemplate::readme_document
---

<!-- README.md and index.md are generated from README.Rmd.
     Edit that file and render it the usual way: rmarkdown::render(), devtools::build_readme(), or the Knit button.
     The cynkratemplate package must be installed; it supplies the output format. -->

```{r setup, include = FALSE}
pkgload::load_all()
```


# RSQLite

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![rcc](https://github.com/r-dbi/RSQLite/workflows/rcc/badge.svg)](https://github.com/r-dbi/RSQLite/actions)
[![Coverage Status](https://codecov.io/gh/r-dbi/RSQLite/graph/badge.svg)](https://app.codecov.io/gh/r-dbi/RSQLite/tree/main)
[![CRAN status](https://www.r-pkg.org/badges/version/RSQLite)](https://cran.r-project.org/package=RSQLite)
[![CII Best Practices](https://bestpractices.coreinfrastructure.org/projects/3234/badge)](https://bestpractices.coreinfrastructure.org/projects/3234)
<!-- badges: end -->

Embeds the SQLite database engine in R, providing a DBI-compliant interface.
[SQLite](https://www.sqlite.org/index.html) is a public-domain, single-user, very light-weight database engine
that implements a decent subset of the SQL 92 standard,
including the core table creation, updating, insertion, and selection operations, plus transaction management.

## Goals and non-goals

RSQLite aims to:

- Embed SQLite in R.
  The engine's sources and the bundled extensions ship with the package, so no external database software is needed.
- Implement the DBI specification for SQLite,
  with the full DBItest suite running as part of the package's own tests.
- Behave the same on every installation,
  by compiling the included SQLite sources into the package rather than linking a SQLite from the system.
  The only system library `configure` looks for is libcurl, and only for the optional HTTP VFS.
- Make the SQLite-specific features that DBI does not cover available,
  such as loadable extensions through `initExtension()`,
  and read-only access to remote database files when libcurl is present at build time.

It is explicitly not trying to:

- Use a SQLite installed on the system.
  The version and the build options are those of the sources included here.
- Define the database interface.
  That is DBI's part, which is why the vignette tells you to load DBI rather than RSQLite.
- Make up for what SQLite does not do.
  It is a single-user, light-weight engine covering a subset of SQL 92, and RSQLite exposes it as it is.
- Keep the pre-DBI API alive.
  Functions such as `dbGetPreparedQuery()`, `sqliteBuildTableDefinition()` and `make.db.names()` remain only as deprecated shims for existing code.

## Installation

You can install the latest released version from CRAN with:

```R
install.packages("RSQLite")
```

Or install the latest development version from GitHub with:

```R
# install.packages("pak")
pak::pak("r-dbi/RSQLite")
```

## Extensions

RSQLite supports loadable extensions.
An experimental HTTP/HTTPS VFS is available when the package is built with libcurl support,
enabling read-only access to remote SQLite databases.
See the help pages for `initExtension()`, `sqliteHttpConfig()`, `sqliteRemote()`, and `sqliteHasHttpVFS()`.

Discussions associated with DBI and related database packages take place on [R-SIG-DB](https://stat.ethz.ch/mailman/listinfo/r-sig-db).
The website [Databases using R](https://db.rstudio.com/) describes the tools and best practices in this ecosystem.

## Basic usage

```{r}
library(DBI)
# Create an ephemeral in-memory RSQLite database
con <- dbConnect(RSQLite::SQLite(), ":memory:")

dbListTables(con)
dbWriteTable(con, "mtcars", mtcars)
dbListTables(con)

dbListFields(con, "mtcars")
dbReadTable(con, "mtcars")

# You can fetch all results:
res <- dbSendQuery(con, "SELECT * FROM mtcars WHERE cyl = 4")
dbFetch(res)
dbClearResult(res)

# Or a chunk at a time
res <- dbSendQuery(con, "SELECT * FROM mtcars WHERE cyl = 4")
while (!dbHasCompleted(res)) {
  chunk <- dbFetch(res, n = 5)
  print(nrow(chunk))
}
# Clear the result
dbClearResult(res)

# Disconnect from the database
dbDisconnect(con)
```

## Acknowledgements

Many thanks to Doug Bates, Seth Falcon, Detlef Groth, Ronggui Huang, Kurt Hornik,
Uwe Ligges, Charles Loboz, Duncan Murdoch, and Brian D. Ripley
for comments, suggestions, bug reports, and/or patches.

---

Please note that the 'RSQLite' project is released with a [Contributor Code of Conduct](https://rsqlite.r-dbi.org/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.
